Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2020-02-06 manuscript/main.Rmd
AGC uncertainty + manuscript tweaks

abstract: |
   Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity accurate long term reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite era (< 1972) is an imperative. Here, we use a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin. Our study leveraged Structure-from-Motion and a novel Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large sub meter resolution orthomosaic (~82800 ha) geo-referenced to ~4.7 $\pm$ 4.3 m. A comparison with contemporary LULC data showed a shift from previously highly regular industrial deforestation of large areas, to discrete smallholder farming clearing, increasing landscape fragmentation but also providing opportunties for substantial forest regrowth. Efforts to quantify long term canopy texture changes and a link to above ground carbon had limited to no success. Our analysis provides methods and insights into key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale, providing a historical context for past and ongoing forest research in the area.
keywords: |
  Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin
acknowledgement: |
  This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE). K.H. acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668). K.H. grateful for the support of T.d.H. volunteering his time in scanning the images.
authorcontributions: |
  K.H. conceived and designed the study, analyzed the data, prepared figures, tables and authored the final draft of the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V and L.W. reviewed the final manuscript.
conflictsofinterest: |
  The authors declare no conflict of interest. The founding sponsors had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, an in the decision to publish the results.
output: rticles::mdpi_article
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  library(orthodrc)
  library(raster)
  library(sf)
  library(rgdal)
  library(ggplot2)
  library(ggthemes)
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(landscapemetrics)

@@

# add a row if forest cover is 100% accurate
if(forest_stats$count[1] == 100){
  if(forest_stats$value == 1){
   forest_stats <- rbind(c(0,0),forest_stats)
  }
}

@@

Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015]. Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global carbon emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function [@fauset2017; @brinck2017; @didham1999]. Strong fragmentation effects **decrease** the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018]. Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].

@@

Historical estimates of Land-Use and Land-Cover (LULC), in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988]. Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post World War II Germany [@nita2018]. Survey data across the African continent is less common, inaccessible or both. Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of **1930s**.

Across the central Congo Basin most of these historical images were collected within the context of national cartographic efforts by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville), DR Congo. Despite the existence of large archives of aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no studies have valorized these data. The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015]. Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].

Here, we use a combination of historical aerial photography **(1958)** and contemporary remote sensing data **(2000-2018)** to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the **anthrophocene** [@lewis2015] with current assessments. Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin. Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.

We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation. Our methodology aims to **provide** a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity. We further contextualize the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin. Our fast scalable mapping approach for historical aerial survey data, using limited supervised input, would further support long-term land-use and land-cover change analysis across the central Congo Basin.

@@

Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa, DR Congo (then Léopoldville). Black-and-White **panchromatic** images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between 09:00 and 11:00 local time. Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negatives of 180 by 180 mm. Flights were flown at an average absolute altitude of ~5200 m above sea level, covering roughly 18 530 km^2^ at an approximate scale of 1/40 000. The use of the integrated autograph system ensured timely acquisition of pictures with a precise overlap (~1/3) between images. This large overlap between images together with flight parameters would allow post-processing, using stereographs, to create accurate topographic maps. Original data from this campaign are stored in the [Royal Museum for Central Africa](https://www.africamuseum.be/en) in Tervuren, Belgium (Figure 1).

```{r echo=FALSE, out.width='75%', fig.align="center", fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the right side margin of the image, such as acquisition time and fleight height."}
knitr::include_graphics('./figures/print_examples.jpg')
```

@@

```{r echo=FALSE, out.width='75%', fig.align="center", message=FALSE, warning=FALSE, fig.cap="Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study. The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km). The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10km). The location of various permanent sampling plots are shown as x, +, and open squares and triangles for the mixed, mono-dominant and edge plots respectively. The grey polygon delineates the current day Yangambi Man and Biosphere reserve. The inset, top right, situates the greater Yangabmi region (white rectangle) with the DR Congo. The full flight plan and details are shown in Appendix Figure 1 and 2"}
knitr::include_graphics('./figures/flight_paths.png')
```

@@

Data was processed into a georeferenced orthomosaic using a Structure-from-motion (SfM, @ullman1979) approach implemented in [Agisoft Metashape](http://www.agisoft.com) version 1.5.2 (Agisoft LLC, St. Petersburg, Russia). An orthomosaic corrects remote sensing data to represent a perfectly downward looking image, free from perspective distortions due to topography and camera tilt. Using the SfM technique features, areas in images with a large degree of similarity, are matched across various images to reconstruct a three dimensional scene (topography) from two-dimensional image sequences. During the SfM analysis we masked clouds, glare or large water bodies such as the Congo river.

We calculated the orthomosaic using a low resolution point cloud and digital elevation map (DEM). Additional ground control points were provided to assist in the referencing of image and constrain the optimization routine used in the SfM algorithm. Ground control points consisted of permanent structures which could be verified in both old and new aerial imagery (i.e. ESRI World Imagery) and consisted of corner points of build structures (e.g. a building, bridge or swimming pool etc.). **Although most clouds were removed during the SfM routine some were retained to provide sufficient SfM tie points to maximize continuous forest coverage in the final orthomosaic.** The final scene was cropped, to provide consistent wall to wall coverage of the reconstructed scene. The orthomosaic was exported as a geotiff for further georeferencing in QGIS [@qgisdevelopmentteam2018] using the georeferencer plugin (version 3.1.9) and additional ESRI World Imagery high resolution reference data. We used **3^rd^** degree polynomial and 16 ground control points to correct the final image. Ground control points, raw image data and final processed image are provided in addition to measures of uncertainty such as mean and median error across all ground control points. All subsequent analysis are executed on the final geo-referenced orthomosaic or subsets of it.

@@

We report the CNN accuracy based upon the IoU of our out-of-sample validation dataset of synthetic landscapes. In addition, we report confusion matrices for all pixels across the homogeneous validation polygons, as well as the training and testing polygons (see Figure 5). Furthermore, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 08:55 GMT or 09:55 local time) to classify and assess the robustness of the CNN algorithm on contemporary remote sensing imagery with similar spectral and spatial characteristics. We used the Global Forest Change version 1.6 (GFC, tile 10N-020E) [@hansen2013] map data, capturing **deforestation** up to 2011, to quantify accuracy on downsampled CNN Geo-Eye classification results. Once more, we report the confusion matrix of between the GFC and CNN derived forest cover maps, masking clouds and cloud shadows. To summarize confusion matrices we report accuracy as:

@@

To map long term LULCC in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013]. Using the GFC data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included GFC pixels which recorded no forest loss throughout the whole 2000 - 2018 period, **as such locations which would see reforestation or deforestation between 2000 and 2018 would be marked as non-forest (i.e. disturbed)**. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30 m GFC resolution using a nearest neighbour approach. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. We provide summary statistics of historical and contemporary **deforestation and reforestation**. We map permanent forest loss after 1958, **reforestation** after 1958, **recent deforestation** and long term (stable) forest cover. **All references to changes over time in the context of our analysis explicitly compare the historical and contemporary periods from hereon forth.**

@@

To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape (fragmentation) metrics [@dale1999]. Landscape metrics provide a mathematical framework for the analysis of discrete land‐cover classes and allows us to capture their composition and configuration. These metrics are therefore commonly used to compare how landscapes change over time [@hesselbarth2019]. In particular, fractals provide a way to quantify complex natural landscapes, including their self-similarity, across scales [@Li2000; @Mandelbrot1977] We report the ratio of edge to area and the fractal dimension to quantify landscape complexity of forest disturbances. A fractal dimension closer to 2 suggest a more complex (fragmented) landscape.

Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package landscapemetrics [@hesselbarth2019]. We provide mean and standard deviation on edge, area, their ratio and fractal dimension for both the historical and contemporary @hansen2013 forest cover maps.

We estimated above ground carbon (AGC) losses and gains, **due to deforestation and reforestation**, using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2). We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data. Unlike standard square 1 ha plots, edge plots were set back 200 m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 1). We further confirmed that forest edge plots did not show a significantly different AGC compared to those of non-edged / mixed forest plots (Mann Whitney U test, p < 0.05). Thus it was not necessary to explicitly quantify changes in AGC caused by edge effects. Moreover, we used the mean values of the mixed forest as representative for potential AGC losses. Despite the challenges inherent in quantifying AGC for forest edges we mapped the total extent of the edges in the contemporary landscape. To align our landscape analysis with exploratory analysis of the survey data we used a buffer of 200 m to estimate the extent of forest edges and patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on age and the crop type. For example, the AGC values varied from 168.67 to 86.55 Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications). These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see Appendix Table 1). We therefore use both the estimates of old-regrowth and mixed forest to estimate AGC for regrowth. We did not have sufficient data to account for individual changes in AGC across plantations.

@@

For both site based and scene analysis we correlated PC values with permanent sample plot inventory data such as stem density, above ground biomass and tree species richness. Additional comparisons are made between contemporary Geo-Eye data and the historical orthomosaic derived PC values. Due to the few available permanent sampling plots in both scenes we used a non-parametric paired signed rank (Wilcoxon) test [@wilcoxon1945] to determine differences between the PC values of the Geo-Eye and historical orthomosaic image scenes across mono-dominant and mixed forest types. In all analysis, mono-dominant site 4 was removed from the analysis due to cloud contamination.

@@

Our analysis provides a first spatially explicity historical composite of aerial survey images in support of mapping land-use and land-cover within the Congo Basin. The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent. The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across ~82800 ha (~23x36 km, Figure 2). The overall accuracy of the SfM orthomosaic composition was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel. The resulting georeferenced scene had a spatial accuracy of approximately 23 m. Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6 m). The orthomosaic served as input for all subsequent LULCC analysis with all derived maps provided with the manuscript repository (see data & code availability statements below).

```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="orthomosaic", fig.cap="Overview of the final orthomosaic of the greater Yangambi region (A) as well as the forest cover classification uncertainty (B) used to generate the final Land Land Cover map (see Figure 6). Homogeneous polygons used in the generation of the synthetic landscape for Convolutional Neural Network training, testing and validation are marked as dashed and full lines for forest and non-forest regions, respectively. Training, testing and validation regions are denoted in panel B in green, red and orange, respectivelly. Black polygon outlines denote cloud and image stitch line regions which were manually excluded from analysis, but retained in calculation of validation statistics (see Table 2)."}
knitr::include_graphics('./figures/orthomosaic_maps.png')
```

@@

The CNN deep learning classifier reached an Intersection-over-Union accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data. Using all pixels within the **validation** polygons (Figure 5) showed a similar accuracy value of ~98%. Using all polygons across the scene, including those used in the generation of testing and training synthetic landscapes, increased the accuracy to ~99% (Table 2). A comparison with recent pan-chromatic Geo-Eye data shows good agreement, with an accuracy of ~87% across all pixels, between the landsat based GFC data and downscaled CNN results (Table 2 and Figure 7).

@@

Scaling our classifier to the whole historical orthomosaic we detected ~16200 ha (or ~20% of the scene) of disturbed forests. A large fraction of the disturbed area was restored in the period between the two acquisition periods. In total, 9918 ha, or little over half of the affected forest was restored (Figure 5C-D, dark blue). Recent deforested areas, as registered through satellite remote sensing **(> 2000)**, approximate 8776 ha (Table 3, Figure 6 - light green).

>>>>

abstract: |
   Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity, accurate long term reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite era (< 1972) is an imperative. Here, we used a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin. Our study leveraged Structure-from-Motion and a Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large orthomosaic (~82 800 ha) geo-referenced to ~4.7 $\pm$ 4.3 m at sub meter resolution. A comparison with contemporary LULC data showed a shift from previously highly regular industrial deforestation of large areas, to discrete smallholder farming clearing, increasing landscape fragmentation but also providing opportunties for substantial forest regrowth. We estimated above ground carbon gains through reforestation to range from 811 to 1592 Gg C, partially offseting historical deforestation (2416 Gg C), in our study area. Efforts to quantify long term canopy texture changes and their link to above ground carbon had limited to no success. Our analysis provides methods and insights into key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale, providing a historical context for past and ongoing forest research in the area.
keywords: |
  Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin
acknowledgement: |
  This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE). K.H. acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668). K.H. grateful for the support of T.d.H. volunteering his time in scanning the images.
authorcontributions: |
  K.H. conceived and designed the study, analyzed the data, prepared figures, tables and authored the final draft of the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V and L.W. reviewed the final manuscript.
conflictsofinterest: |
  The authors declare no conflict of interest. The founding sponsors had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, an in the decision to publish the results.
output: rticles::mdpi_article
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  library(raster)
  library(sf)
  library(rgdal)
  library(ggplot2)
  library(ggthemes)
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(landscapemetrics)

@@

# add a row if forest cover is 100% accurate
# messes with table otherwise
if(forest_stats$count[1] == 100){
  if(forest_stats$value == 1){
   forest_stats <- rbind(c(0,0),forest_stats)
  }
}

@@

Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015]. Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global carbon emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function [@fauset2017; @brinck2017; @didham1999]. Strong fragmentation effects decrease the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018]. Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].

@@

Historical estimates of Land-Use and Land-Cover (LULC), in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988]. Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post World War II Germany [@nita2018]. Survey data across the African continent is less common, inaccessible or both. Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930s.

Across the central Congo Basin most of these historical images were collected within the context of national cartographic efforts by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville), DR Congo. Despite the existence of large archives of aerial survey imagery of African rainforest (Figure 1, Appendix Figure 3), as of yet, no studies have valorized these data. The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015]. Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].

Here, we use a combination of historical aerial photography (1958) and contemporary remote sensing data (2000-2018) to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the anthrophocene [@lewis2015] with current assessments. Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin. Past research in the region allows for thorough assessment of LULCC from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.

We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation. Our methodology aims to provide a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity. We further contextualize the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin. Our fast scalable mapping approach for historical aerial survey data, using limited supervised input, would further support long-term land-use and land-cover change analysis across the central Congo Basin.

@@

Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1 - 2 and Appendix Table 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa, DR Congo (then Léopoldville). Black-and-White panchromatic images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between 09:00 and 11:00 local time. Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negatives of 180 by 180 mm. Flights were flown at an average absolute altitude of ~5200 m above sea level, covering roughly 18 530 km^2^ at an approximate scale of 1/40 000. The use of the integrated autograph system ensured timely acquisition of pictures with a precise overlap (~1/3) between images. This large overlap between images together with flight parameters would allow post-processing, using stereographs, to create accurate topographic maps. Original data from this campaign are stored in the [Royal Museum for Central Africa](https://www.africamuseum.be/en) in Tervuren, Belgium (Figure 1, Appendix Figure 3).

```{r echo=FALSE, out.width='75%', fig.align="center", fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the right side margin of the image, such as acquisition time and flight height."}
knitr::include_graphics('./figures/print_examples.jpg')
```

@@

```{r echo=FALSE, out.width='75%', fig.align="center", message=FALSE, warning=FALSE, fig.cap="Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study. The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km). The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10 km). The location of various permanent sampling plots are shown as x, +, and open squares and triangles for the mixed, mono-dominant and edge plots respectively. The grey polygon delineates the current day Yangambi Man and Biosphere reserve. The inset, top right, situates the greater Yangabmi region (white rectangle) with the DR Congo. The full flight plan and details are shown in Appendix Figure 1 and 2"}
knitr::include_graphics('./figures/flight_paths.png')
```

@@

Data was processed into a georeferenced orthomosaic using a Structure-from-motion (SfM, @ullman1979) approach implemented in [Agisoft Metashape](http://www.agisoft.com) version 1.5.2 (Agisoft LLC, St. Petersburg, Russia). An orthomosaic corrects remote sensing data to represent a perfectly downward looking image, free from perspective distortions due to topography and camera tilt. Using the SfM technique features, areas in images with a large degree of similarity, are matched across various images to reconstruct a three dimensional scene (topography) from two-dimensional image sequences. During the SfM analysis we masked most clouds, glare or large water bodies such as the Congo river.

We calculated the orthomosaic using a low resolution point cloud and digital elevation map (DEM). Additional ground control points were provided to assist in the referencing of image and constrain the optimization routine used in the SfM algorithm. Ground control points consisted of permanent structures which could be verified in both old and new aerial imagery (i.e. ESRI World Imagery) and consisted of corner points of build structures (e.g. a building, bridge or swimming pool etc.). Although most clouds were removed during the SfM routine some were retained to provide sufficient SfM tie points to maximize continuous forest coverage in the final orthomosaic. The final scene was cropped, to provide consistent wall-to-wall coverage of the reconstructed scene. The orthomosaic was exported as a geotiff for further georeferencing in QGIS [@qgisdevelopmentteam2018] using the georeferencer plugin (version 3.1.9) and additional ESRI World Imagery high resolution reference data. We used 3^rd^ degree polynomial and 16 ground control points to correct the final image. Ground control points, raw image data and final processed image are provided in addition to measures of uncertainty such as mean and median error across all ground control points. All subsequent analysis are executed on the final geo-referenced orthomosaic or subsets of it.

@@

We report the CNN accuracy based upon the IoU of our out-of-sample validation dataset of synthetic landscapes. In addition, we report confusion matrices for all pixels across the homogeneous validation polygons, as well as the training and testing polygons (see Figure 5). Furthermore, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 08:55 GMT or 09:55 local time) to classify and assess the robustness of the CNN algorithm on contemporary remote sensing imagery with similar spectral and spatial characteristics. We used the Global Forest Change version 1.6 (GFC, tile 10N-020E) [@hansen2013] map data, capturing deforestation up to 2011, to quantify accuracy on downsampled CNN Geo-Eye classification results. Once more, we report the confusion matrix between the GFC and CNN derived forest cover maps, masking clouds and cloud shadows. To summarize confusion matrices we report accuracy as:

@@

To map long term LULCC in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) (lossyear) map data [@hansen2013]. Using the GFC data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included GFC pixels which recorded no forest loss throughout the whole 2000 - 2018 period. Forest loss in the context of GFC is defined as "a stand-replacement disturbance, or a change from a forest to non-forest state". As such, locations which would see reforestation or deforestation between 2000 and 2018 would be marked as non-forest (i.e. disturbed). As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30 m GFC resolution using a nearest neighbour approach. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. We provide summary statistics of historical and contemporary deforestation and reforestation. We map permanent deforestation after 1958, reforestation after 1958, recent deforestation and long term (stable) forest cover. All references to changes over time in the context of our analysis explicitly compare the historical and contemporary periods from hereon forth.

@@

To quantify changes in the structure of forest cover and its disturbances we used spatial landscape pattern analysis (i.e. fragmentation) metrics [@dale1999]. Landscape metrics provide a mathematical framework for the analysis of discrete land‐cover classes and allows us to capture their composition and configuration. These metrics are therefore commonly used to compare how landscapes change over time [@hesselbarth2019]. In particular, fractals provide a way to quantify complex natural landscapes, including their self-similarity across scales [@Li2000; @Mandelbrot1977]. We report the ratio of edge to area and the fractal dimension to quantify landscape complexity of forest disturbances. A fractal dimension closer to 2 suggest a more complex (fragmented) landscape.

Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package "landscapemetrics" [@hesselbarth2019]. We provide mean and standard deviation on edge, area, their ratio and fractal dimension for both the historical and contemporary @hansen2013 forest cover maps.

We estimated above ground carbon (AGC) losses and gains, due to deforestation and reforestation, using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2). We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data. Unlike standard square 1 ha plots, edge plots (163.03 $\pm$ 19.39 Mg C ha^-1^, N = 5) were set back 200 m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 2). We further confirmed that forest edge plots, as compared to mixed forest plots (160.48 $\pm$ 23.84 Mg C ha^-1^, N = 8, see Appendix Table 3 for all forest types), did not show a significantly different AGC (Mann Whitney U test, p < 0.05). Thus it was not necessary to explicitly quantify changes in AGC caused by edge effects. Moreover, we used the mean value, and its uncertainty (i.e. standard deviation), of the mixed forest as representative for potential AGC losses. Despite the challenges inherent in quantifying AGC for forest edges we mapped the total extent of the edges in the contemporary landscape. To align our landscape analysis with exploratory analysis of the survey data we used a buffer of 200 m to estimate the extent of forest edges and patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on age and the crop type. For example, the AGC values varied from 86.55 to 168.67 Mg C ha^-1^, for *Elaeis guineensis* (oil palm) and *Hevea brasiliensis* (rubber tree) plots respectively (@bustillo2018, personal communications). These higher values are in line with the mixed plot AGC estimates (160.48 $\pm$ 23.84 Mg C ha^-1^, N = 8) in the area, while the palm plantations resemble old-regrowth values (81.87 Mg C ha^-1^, N = 1). To quantify AGC in reforested areas we therefore use both AGC estimates of old-regrowth and mixed forest, as lower and upper bounds. We did not have sufficient data to account for individual changes in AGC across plantations.

@@

For both site based and scene analysis we correlated PC values with permanent sample plot inventory data such as stem density, above ground biomass and tree species richness (Appendix Table 2-3). Additional comparisons are made between contemporary Geo-Eye data and the historical orthomosaic derived PC values. Due to the few available permanent sampling plots in both scenes we used a non-parametric paired signed rank (Wilcoxon) test [@wilcoxon1945] to determine differences between the PC values of the Geo-Eye and historical orthomosaic image scenes across mono-dominant and mixed forest types. In all analysis, mono-dominant site 4 was removed from the analysis due to cloud contamination.

@@

Our analysis provides a first spatially explicity historical composite of aerial survey images in support of mapping land-use and land-cover within the Congo Basin. The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent. The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across ~82 800 ha with a resolution of 0.88 m / pixel (or ~23x36 km, Figure 2). The overall spatial accuracy of the SfM orthomosaic composition using the sparse cloud DEM (with a resolution of 45.8 m/pixel) was limited to approximately 23 m. Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6 m). The orthomosaic served as input for all subsequent LULCC analysis with all derived maps provided with the manuscript repository (see data & code availability statements below).

```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="orthomosaic", fig.cap="Overview of the final orthomosaic of the greater Yangambi region (A) as well as the forest cover classification uncertainty (B) used to generate the final Land Cover map (see Figure 6). Homogeneous polygons used in the generation of the synthetic landscape for Convolutional Neural Network training, testing and validation are marked as dashed and full lines for forest and non-forest regions, respectively. Training, testing and validation regions are denoted in panel B in green, red and orange, respectivelly. Black polygon outlines denote cloud and image stitch line regions which were manually excluded from analysis, but retained in calculation of validation statistics (see Table 2)."}
knitr::include_graphics('./figures/orthomosaic_maps.png')
```

@@

The CNN deep learning classifier reached an Intersection-over-Union accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data. Using all pixels within the validation polygons (Figure 5) showed a similar accuracy value of ~98%. Using all polygons across the scene, including those used in the generation of testing and training synthetic landscapes, increased the accuracy to ~99% (Table 2). A comparison with recent pan-chromatic Geo-Eye data shows good agreement, with an accuracy of ~87% across all pixels, between the landsat based GFC data and downscaled CNN results (Table 2 and Figure 7).

@@

Scaling our classifier to the whole historical orthomosaic we detected ~16 200 ha (or ~20% of the scene) of disturbed forests. A large fraction of the disturbed area was restored in the period between the two acquisition periods. In total, 9918 ha, or little over half of the affected forest was restored (Figure 5C-D, dark blue). Recent deforested areas, as registered through satellite remote sensing (> 2000), approximate 8776 ha (Table 3, Figure 6 - light green).
